---
title: "01_load"
author: "Group_23"
format:
  html:
    embed-resources: true
editor: visual
---

# 01. Load the data

```{r}
library(tidyverse)
```

## Loading the raw data from `/data/_raw` directory

Raw data has to be downloaded to `/data/_raw` directory from [Kaggle](https://www.kaggle.com/datasets/piotrgrabo/breastcancerproteomes/data/). The data could not be retrieved automatically from the server, and the data will not be loaded in the project repository on GitHub. If you are executing this end-to-end be sure that the `77_cancer_proteomes_CPTAC_itraq.csv` is unzipped in your own \_raw directory.

```{r}
#| eval: true 
PAM50 <- read.csv("../data/_raw/PAM50_proteins.csv")
proteomes_77 <- read.csv("../data/_raw/77_cancer_proteomes_CPTAC_itraq.csv")
clinical <- read.csv("../data/_raw/clinical_data_breast_cancer.csv")
```

## Save data as .tsv files in `/data` directory

```{r}
#| eval: true 
PAM50 |> 
  write_tsv(file = "../data/PAM50_proteins.tsv")

proteomes_77 |> 
  write_tsv(file = "../data/77_cancer_proteomes_CPTAC_itraq.tsv")       

clinical |> 
  write_tsv(file = "../data/clinical_data_breast_cancer.tsv")    

```
